---
title: "Not Just R: Part 1"
description: |
  Putting the spotlight on Python.
author:
  - name: Juan David Acosta
date: 2022-04-30
categories:
  - Python
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

From my undergrad coursework, as well as through some online courses, I have worked at building up my R proficiency, to the point where I am very comfortable using it as my go-to programming language for analysis and basic programming. Virtually all the assignments I did for my undergrad courses that are shared on this website were written using R code.

However, R is not the only programming language I know. I have also been exposed to Python programming, both through some undergrad courses and through online courses. But, since I tend to conduct analyses using R, I don't get to practice using my Python skills as much. I am hoping to change that through both this blog and this website as a whole. Like I am doing with some blog posts where I set out a coding example using R, I would also like to do the same using Python. And that is made easier by the fact that Python code can be written and executed using RStudio, even alongside R code if you wish. I am also planning on completing some personal projects using Python, but that is still in the planning stages.

Basically, even though R is the standard when it comes to programming languages with a statistical analysis focus, Python has also shone in the data science world, and it is vital to continue to develop proficiency in both programming languages. So keep on the lookout for an upcoming Part 2 where I introduce a Python coding example!

But, as a little teaser, and in homage to what most, if not all, Python beginners experience when starting to learn the language, we will print out the phrase "Hello World!"

In order to run Python code in RStudio, aside from already having a local installation of Python on your machine (I use Anaconda myself), you need the *Reticulate* R package. If you don't have it already installed, please do so before attempting to run the following code.

```{r, echo=TRUE}
library(reticulate)
```

Now, we can write Python code as if we were in any other Python IDE! Let's print "Hello World!"

```{python, echo=TRUE}
print("Hello World!")
```

We did it! That's just a tiny taste of the Python code we'll go through in future installments of "Not Just R". Next time we'll do some calculations, and maybe even explore Python libraries, which are equivalent to R packages in the sense that they extend the capabilities of base Python, sometimes for more specialized topics.

That's a wrap on today's post!
